Think deep for reef resilience

### 2025-09-25 paper/paper_draft.qmd
Minor tweaks to wording

Climate change is impacting species survival, persistence and ecosystem functions globally.
This is particularly true for coral reefs, which are vulnerable to increasing magnitude and
variation in environmental stressors such as ocean temperature, acidification and nutrient
loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events driven by climate
change can cause reef-building coral population declines [@hughes2017; @hughes2018;
@bozec2025], possibly leading to long-term and large-scale coral-macroalgae regime shifts
[@arif2022; @wild2011]. Climate impacts on coral reefs are decreasing the ability of coral
reefs to provide ecological functions such as benthic habitat and nutrient cycling
[@forster2021; @williams2019] as well as ecosystem services such as food, protection,
tourism and Indigenous cultural value [@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes. Coral
reef ecosystem resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survive thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third, the
maintenance, and accretion, of the carbonate framework that provides reef structure
[@perry2012]. We aim to assess how these mechanisms interact under projected climate
conditions to identify the characteristics of resilient reefs through scenario-based coral
population modelling for the Great Barrier Reef ecosystem.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest as
the predictors, including the median reef depth, mean DHW for the given GCM, log10-weighted
incoming connectivity, log10-weighted outgoing connectivity, and log10-outgoing-to-incoming
connectivity ratio. The full dataset consisted of the time series cluster assignment of each
of the 2197 reefs under the five GCMs explored, for a total of 10,985 samples. The dataset
\*\*\[\] was randomly split into training (60%) and testing (40%) subsets to ensure model
generalisability. The RF classifier achieved 80% accuracy on the training set and 59% on the
test set.

@@

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically (one-at-a-time) varied across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so, are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and incoming connectivity, two-way partial dependence
analysis was performed on grids along the range of these two features.

>>>>

Climate change is impacting species survival, persistence and ecosystem functions globally.
This is particularly true for coral reefs, which are vulnerable to increasing magnitude and
variation in environmental stressors such as ocean temperature [@foster2021; @gattuso2014].
Mass bleaching and associated mortality events driven by climate change can cause
reef-building coral population declines [@hughes2017; @hughes2018; @bozec2025], possibly
leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011].
Climate impacts on coral reefs are decreasing the ability of coral reefs to provide
ecological functions such as benthic habitat and nutrient cycling [@forster2021;
@williams2019] as well as ecosystem services such as food, protection, tourism and
Indigenous cultural value [@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes. Coral
reef ecosystem resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - is conceptualised to depend on three interconnected mechanisms. First, the
ability of individual corals to survive thermal stress. Second, the connectivity networks
that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third,
the maintenance, and accretion, of the carbonate framework that provides reef structure
[@perry2012]. We aim to assess how these mechanisms interact under projected climate
conditions to identify the characteristics of resilient reefs through scenario-based coral
population modelling for the Great Barrier Reef ecosystem.

@@

Partial Dependence Analysis was conducted to understand the relationship between features
and bioregion-scale time series cluster predictions using a Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the estimated median reef depth, mean DHW for the given GCM,
log10-weighted incoming connectivity, log10-weighted outgoing connectivity, and
log10-outgoing-to-incoming connectivity ratio. The full dataset consisted of the time series
cluster assignment of each of the 2197 reefs under the five GCMs explored, for a total of
10,985 samples. The dataset was randomly split into training (60%) and testing (40%) subsets
to train and assess classification performance. The RF classifier achieved 80% accuracy on
the training set and 60% on the test set.

@@

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically varied (one-at-a-time) across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. Use of the median predicted
probability was also assessed but was found not to substantially influence the
interpretation of results (and so are not shown here). The process isolates the marginal
effect of each predictor by averaging over the natural distribution of all other
representative reef characteristics in the test dataset. Additionally, to assess possible
interaction between reef depth and incoming connectivity, two-way partial dependence
analysis was performed on grids along the range of these two features.
